Making the Best of a Bad Situation

### 2022-03-22 paper.tex
nits and changes

Much of the time, this burden can be mitigated by automation, such as when the data structure being generated
is composed entirely of structures that already have generators, or $P$ is described as an inductive
relation \cite{gggir}. However, the case of inferring generators for \emph{numerical} properties---those involving constriants over numbers--- is not well
handled by prior work. Numerical properties appear regularly in
systems programming tasks \cite{ironfleet} where structured data described by inductive invariants is rare, inputs regularly
take the form of pointers and integers, and many function preconditions amount to bounds-checking or ensuring
that certain fields in different structures are related in some way.

In this extended abstract, we develop a method for inferring random generators for a small class of numerical properties akin
to the kind commonly found in systems programming. Because manual generator-writing for numerical properties is especially
tedious, our goal is to infer generators that are comparable to ones that a user might write as a first cut.
The method proceeds by developing a number of generator ``candidates'', expressed in a DSL we call \texttt{ALuck}, described in \autoref{sec:luck}.
These candidates mimic the way that a human might write a generator for numerical preconditions: by randomly choosing values
one variable at a time, constraining subsequent choices by the previous values chosen. In \autoref{sec:sci}, we show how candidates
are chosen from the space of possible \texttt{ALuck} generators.
As we will see, it is sometimes the case that \emph{none} of the candidates are all that good. In this case, we must make the best
of a bad situation, and discover the best candidate among the set. A learning-theoretic method for this is presented in \autoref{sec:bandits}.
Finally, in \autoref{sec:impl}, we describe our implementation of
this algorithm, present benchmarks, and briefly discuss future work.

@@

The syntax of propositions that we handle are shown in
\autoref{fig:precond-syntax-and-aluck}. These are conjunctions and negations of
arithmetical equalities and inequalities with variables, which range over
the arguments of the function under test. The allowable expressions in these inequalities are
essentially multi-linear functions, where each variable occurs with degree at
most one. While this form is restrictive, we have found empirically that this covers
a wide range of preconditions in systems verification. Moreover, there
appears to be no inherent difficulty to extending the technique to handle
other numerical operators (div, mod), and constraints over structured types like lists: we present the algorithm as-is to
simplify the discussion.

@@

To make the generator inference problem tractable, we fix the
syntactic form of the generators. Our generators are written
in a language called \texttt{ALuck} (for \textit{arithmetic} \texttt{Luck})
inspired by \texttt{Luck} \cite{luck}, a domain-specific language for writing
generators. Generators written \texttt{ALuck} run by sequentially \textit{constraining} and
\textit{concretizing} variables. Every variable in an \texttt{ALuck} generator
begins as a symbolic variable. Constraints over these variables are then added.
Variables can then be concretized, wherein they are replaced by a random value
satisfying the constraints on that variable that have been accumulated thusfar.
The final result of the generator is a map from variables to their (randomly)
chosen values.

@@

Because of this, to infer generators for a property $P$, it suffices to
generate orderings of its variables. Unfortunately, for a property $P$ with
$n$ free variables, there are $n!$ such orderings: a very large search space.
Fortunately, we can prune this search space by only looking for ``relevant"
orderings. If some variables are not related to others, the well-concretization
condition won't care the relative order in which they're generated.

@@

To operationalize this insight, we build a graph $G(P)$ whose nodes are variables
with an edge $(x,y)$ when $x$ and $y$ both occur in one of the conjuncts of $P$. In
this graph, ``unrelated'' variables live in different connected components.
Then, to generate a concretization ordering, we depth-first search $G(P)$, randomly
choosing the next neighbor to explore,
and list variables in the order that they're visited in the graph.
Since different connected components are listed separately, unrelated variables
will not 
The fact that related variables lie in separate connected components means
that 
% we create an undirected graph from a proposition to
% encode the important inter-relationships that a concretization order should
% capture.

@@

To generate our set of generator candidates, we repeatedly run this function.
This may give repeated generators, and so we filter the result for uniqueness.
The number of generators in our set requires a careful
balance. Too few and the set may not contain a generator which succeeds often
enough to rapidly generate our desired number of unique inputs. Too many and the
learning algorithm will converge too slowly to the best generator in the set.
Empirically, we have found that $n^2$ (where $n$ is the number of variables in
$P$) is a number of generators to take.

\section{Generator Learning with Bandits Algorithms}
\label{sec:bandits}
With our bag of generators in hand, we now need to find the best one. The
approach we take will be inspired by the Multi-Armed Bandits
\cite{gittins1979bandit} problem from the theory of reinforcement learning. In short, the multi armed
bandits problem describes a game where an algorithm is repeatedly presented
a fixed set of choices. Each choice gives a different (random) reward, and the
goal of the game is to maximize the total reward. Solving this optimization problem
online is too difficult in an adversarial setting, so algorithms for the multi-armed
bandits problem aim to instead minimize \emph{regret}: how much worse their total reward
was than the best possible total reward in hindsight. To achieve this, bandits algorithms
learn which actions have historically given more reward, and play those more frequently.
In our setting, the ``choices" are our
generator candidates, and the rewards are given by success or failure of a
generator to yield a value. Under this analogy, an algorithm for the Multi-Armed
Bandits problem will let us learn which generators give the best results
\textit{while simultaneously} generating a stream of valid inputs for the
function.

The algorithm we will use is called UCB1 \cite{auer2002finite}. While more sophisticated
bandits algorithms exist, this one is sufficient for our purposes. When given a list of generators
$g_1,\dots g_k$ and a number of rounds $T$ to run for, UCB1 runs the generators $g_i$ in a way that attempts
to learn which generator succeds the most frequently, while also attempting to not waste time by running
failing generators too much. In the end, UCB1 emits the $m \leq T$ successfully-generated values over the course of its run.
The main upshot is that the ratio success ratio $\frac{m}{T}$ of this derived generator should, in the limit,
be no worse than the \emph{best} generator the algorithm was given. 
% UCB1 achieves a
% regret bound of $O(\sqrt{KT\log K})$. While algorithms are known which achieve
% $O(\sqrt{KT})$ regret, UCB1 is simple to implement and understand, and performs
% more than well enough for our setting. In \autoref{alg:ucb1}, we present the
% UCB1 algorithm, modified for use in property-based testing. The algorithm takes
% generator candidates $g_1,\dots g_k$, the property $P$, a number of rounds to
% run for $T$, and returns a list of generated values satisfying $P$.
\begin{restatable}{thm}{regretthm}
Fix a property $P$ and generators $g_1,\dots,g_K$, where $g_i$ succeeds with probability at least $p_i$. Then,
\vspace{-2px}
\[
\lim_{T \to \infty} \mathbb{E}\left[\frac{1}{T}\left|\texttt{ucb1}(g_1,\dots,g_K,P,T)\right|\right] \geq \max_i p_i
\]
\end{restatable}
% Intuitively, this theorem justifies our intuition that the ucb1 algorithm
% ``finds the best generator". As the number of rounds gets large, the ratio of
% the number of valid inputs generated by \texttt{ucb1} to the number of rounds is
% no worse than the success probability of the best generator. 
In other words, the stream of values
emitted by an ``infinite run" of UCB1 acts like a
generator for $P$ which succeeds with probability at least $\max_i p_i$.

@@

Dividing through by $T$, using the regret bound for UCB1, and the fact that $p_i \geq \mu_i$, we have that:
\begin{align*}
\mathbb{E}\left[\frac{1}{T}\sum_{t=1}^T X_{A(t),t}\right] &= \mu_\star - \frac{R(A)}{T}\\
&\geq \mu_{\star} - O\left(\sqrt{\frac{K\log K}{T}}\right)\\
&\geq \max_i p_i - O\left(\sqrt{\frac{K\log K}{T}}\right)
\end{align*}
which approaches $\max_i p_i$ as $T \to \infty$.
\end{proof}

>>>>

Much of the time, this burden can be mitigated by automation, such as when the data structure being generated
is composed entirely of structures that already have generators, or $P$ is described as an inductive
relation \cite{gggir}. However, the case of inferring generators for \emph{numerical} properties---those involving constriants over numbers--- is not well
handled by prior work. Numerical properties appear regularly in
systems programming tasks \cite{ironfleet} where structured data described by inductive invariants is rare, inputs regularly
take the form of pointers and integers, and many function preconditions amount to bounds-checking or ensuring
that certain fields in different inputs are equal.

In this extended abstract, we develop a method for inferring random generators for a small class of numerical properties akin
to the kind commonly found in systems programming. Because manual generator-writing for numerical properties is especially
tedious, our goal is to infer generators that are comparable to ones that a user might write as a first cut.
The method proceeds by developing a number of generator ``candidates'', expressed in a DSL we call \texttt{ALuck}, described in \autoref{sec:luck}.
These candidates mimic the way that a human might write a generator for numerical preconditions: by randomly choosing values
one variable at a time, constraining subsequent choices by the previous values chosen. In \autoref{sec:sci}, we show how candidates
are chosen from the space of possible \texttt{ALuck} generators.
As we will see, it is sometimes the case that \emph{none} of the candidates are all that good. In this case, we must make the best
of a bad situation, and discover the best candidate among the set. A method for this based on the multi-armed bandits problem \cite{banditbook} is presented in \autoref{sec:bandits}.
Finally, in \autoref{sec:impl}, we describe our implementation of
this algorithm and present benchmarks.

@@

The syntax of propositions that we handle are shown in
\autoref{fig:precond-syntax-and-aluck}. The variables range over
the arguments of the function under test. The allowable expressions in these inequalities are
essentially multi-linear functions, where each variable occurs with degree at
most one. While this form is restrictive, we have found empirically that this covers
a wide range of preconditions in systems verification. Moreover, there
appears to be no inherent difficulty to extending the technique to handle
other numerical operators (div, mod), and constraints over structured types like lists: we present the algorithm as-is to
simplify the discussion.

@@

To make the generator inference problem tractable, we fix the
syntactic form of the generators. Our generators are written
in a language called \texttt{ALuck} (for \textit{arithmetic} \texttt{Luck})
inspired by \texttt{Luck} \cite{luck}. Generators written \texttt{ALuck} run by sequentially \textit{constraining} and
\textit{concretizing} variables. Every variable in an \texttt{ALuck} generator
begins as a symbolic variable. Constraints over these variables are then added.
Variables can then be concretized, wherein they are replaced by a random value
satisfying the constraints on that variable that have been accumulated thusfar.
The final result of the generator is a map from variables to their (randomly)
chosen values.

@@

Because of this, to infer generators for a property $P$, it suffices to
generate orderings of its variables. Unfortunately, for a property $P$ with
$n$ free variables, there are $n!$ such orderings.
We can prune this search space by only looking for ``relevant"
orderings. If some variables are not related to others, the well-concretization
condition won't care about the relative order in which they're generated.

@@

To operationalize this insight, we build a graph $G(P)$ whose nodes are variables
with an edge $(x,y)$ when $x$ and $y$ both occur in one of the conjuncts of $P$. In
this graph, ``unrelated'' variables live in different connected components.
Then, to generate a concretization ordering, we depth-first search $G(P)$, randomly
choosing the next neighbor to explore,
and list variables in the order that they're visited in the graph.
Since different connected components are listed separately, concretizations of unrelated variables
will not occur together.
% The fact that related variables lie in separate connected components means
% that 
% we create an undirected graph from a proposition to
% encode the important inter-relationships that a concretization order should
% capture.

@@

To generate our set of generator candidates, we repeatedly run this random DFS procedure.
This may give repeated generators, and so we filter the result for uniqueness.
The number of generators in our set requires a careful
balance. Too few and the set may not contain a generator which succeeds often
enough to rapidly generate our desired number of unique inputs. Too many and the
learning algorithm will converge too slowly to the best generator in the set.
Empirically, we have found that $n^2$ (where $n$ is the number of variables in
$P$) is a number of generators to take.

\section{Generator Learning with Bandits Algorithms}
\label{sec:bandits}
With our bag of generators in hand, we now need to find the best one. The
approach we take will be inspired by the Multi-Armed Bandits
\cite{gittins1979bandit} problem from the theory of reinforcement learning. In short, the Multi-Armed
Bandits problem describes a game where an algorithm is repeatedly presented
a fixed set of choices. Each choice gives a different (random) reward, and the
goal of the game is to maximize the total reward. Solving this optimization problem
online is too difficult in an adversarial setting, so algorithms for the multi-armed
bandits problem aim to instead minimize \emph{regret}: how much worse their total reward
was than the total reward from the best single choice in hindsight. To achieve this, bandits algorithms
learn which actions have historically given more reward, and play those more frequently.
In our setting, the ``choices" are our
generator candidates, and the rewards are given by success or failure of a
generator to yield a value. Under this analogy, an algorithm for the Multi-Armed
Bandits problem will let us learn which generators give the best results
\textit{while simultaneously} generating a stream of valid inputs for the
function.

The algorithm we will use is called UCB1 \cite{auer2002finite}. While more sophisticated
bandits algorithms exist, this one is sufficient for our purposes. When given a list of generators
$g_1,\dots g_k$ and a number of rounds $T$ to run for, UCB1 runs the generators $g_i$ in a way that attempts
to learn which generator succeds the most frequently, while also attempting to not waste time by running
failing generators too much. In the end, UCB1 emits the $m \leq T$ successfully-generated values over the course of its run.
The main upshot is that the ratio success ratio $\frac{m}{T}$ of this derived generator should, in the limit,
be no worse than the \emph{best} generator the algorithm was given. 
% UCB1 achieves a
% regret bound of $O(\sqrt{KT\log K})$. While algorithms are known which achieve
% $O(\sqrt{KT})$ regret, UCB1 is simple to implement and understand, and performs
% more than well enough for our setting. In \autoref{alg:ucb1}, we present the
% UCB1 algorithm, modified for use in property-based testing. The algorithm takes
% generator candidates $g_1,\dots g_k$, the property $P$, a number of rounds to
% run for $T$, and returns a list of generated values satisfying $P$.
\begin{restatable}{thm}{regretthm}
Fix a property $P$ and generators $g_1,\dots,g_K$, where $g_i$ succeeds with probability at least $p_i$. Then,
\vspace{-2px}
\[
\lim_{T \to \infty} \mathbb{E}\left[\frac{1}{T}\left|\texttt{ucb1}(g_1,\dots,g_K,P,T)\right|\right] \geq \max_i p_i
\]
\end{restatable}
% Intuitively, this theorem justifies our intuition that the ucb1 algorithm
% ``finds the best generator". As the number of rounds gets large, the ratio of
% the number of valid inputs generated by \texttt{ucb1} to the number of rounds is
% no worse than the success probability of the best generator. 
In other words, the stream of values
emitted by an ``infinite run" of UCB1 acts like\footnote{In theory, the difference $\left|\frac{1}{T}\mathbb{E}\left[\left|\texttt{ucb1}\right|\right] - \max_i p_i \right|$ converges like $O\left(\frac{\log T}{T}\right)$ \cite[Theorem 1]{auer2002finite}. Empirically, we have found this to be quite rapid.} a
generator for $P$ which succeeds with probability at least $\max_i p_i$.

@@

The regret bound for UCB1 \cite[Theorem 7.2]{banditbook} states that $R(A) \in O\left(\sqrt{KT\log T}\right)$.

Then, dividing through by $T$, using the regret bound for UCB1, and the fact that $p_i \geq \mu_i$, we have that:
\begin{align*}
\mathbb{E}\left[\frac{1}{T}\sum_{t=1}^T X_{A(t),t}\right] &= \mu_\star - \frac{R(A)}{T}\\
&\geq \mu_{\star} - O\left(\sqrt{\frac{K\log T}{T}}\right)\\
&\geq \max_i p_i - O\left(\sqrt{\frac{K\log T}{T}}\right)
\end{align*}
which approaches $\max_i p_i$ as $T \to \infty$.
\end{proof}
